---
title: "Final Project"
author: "Elena Novikova"
date: "5/04/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



```{r}

library(tidyverse)
library(lubridate)
library(magrittr)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(openxlsx)
library(knitr)
library(rvest)
library(choroplethr)
library(choroplethrMaps)

```

&nbsp;
&nbsp;
&nbsp;
&nbsp;

I selected a dataset of Community Health Status Indicators from the Data.gov website. Let's read the description of provided data from the website:

&nbsp;
&nbsp;

```{r Web scraping from data.gov}

# Extracting text from the website:
CHSIdescrip <- "https://catalog.data.gov/dataset/community-health-status-indicators-chsi-to-combat-obesity-heart-disease-and-cancer" %>%
  read_html() %>% html_nodes(".module-content") %>%
  html_text()
CHSIdescrip <- tibble(text = CHSIdescrip)
CHSIdescrip <- CHSIdescrip[[1,1]]   # the text we need is in the first element of the resulting tibble

# using regular expressions to clean up line breaks, tabs:
CHSIheader <- str_extract(CHSIdescrip, "[^\n \t].*")  # extracting the title of the text
CHSIdate <- str_extract(CHSIdescrip, "M.*")           # extracting the date of the last update
CHSItext <- str_extract(CHSIdescrip, "\n +C.*")       # extracting the text of the description
CHSItext <- str_remove_all(CHSItext, "\n  +")         # removing excessive line breaks
CHSItable <- tibble(x = c(CHSIheader, CHSIdate, CHSItext)) # writing the above as observations of a tibble
kable(CHSItable, caption = "CHSItable")  # printing out as a tibble
```

&nbsp;
&nbsp;

The file called DATAELEMENTDESCRIPTION.csv contains column names and column descriptions for each file from the source. I will import it and nest it into a dataframe with values in the first column corresponding to file names, and second column containing dataframes with column descriptions for each file.

&nbsp;
&nbsp;

Here is the list of files in the imported CHSI dataset:

&nbsp;
&nbsp;

```{r working with DATAELEMENTDESCRIPTION file}

DATAELEMENTDESCRIPTION <- read.csv("DATAELEMENTDESCRIPTION.csv") # importing a csv file into a tibble

DATAELEMENTDESCRIPTION <- DATAELEMENTDESCRIPTION %>% # grouping by names of the files in column 1 and nesting
  group_by(PAGE_NAME) %>% 
  nest()
#glimpse(DATAELEMENTDESCRIPTION)

print(DATAELEMENTDESCRIPTION)

```

&nbsp;
&nbsp;

Let's start with looking at column descriptions in the Demographics file:

```{r glimpse at Demographics }
# printing a head of the tibble located in line 1 column 2 of DATAELEMENTDESCRIPTION df
kable(head(DATAELEMENTDESCRIPTION[[1,2]]), caption = "Demographics")
```

The CHSI dataset is labeling counties with Strata IDs. Here is the information provided about stratas, or “Peer County Groups”, on the countyhealthrankings.org website. I used the description given on sheet 1 of the csv file posted on this website:

&nbsp;
&nbsp;

```{r Peer counties decription}
county_fips_descrip <- read_excel("CHSIpeers.xlsx", sheet = 1) # reading the excel file, sheet 1
county_fips_descrip <- county_fips_descrip[[2,1]]             # extracting the text from column 1, line 2
kable(county_fips_descrip, caption = "Peer Counties")         # printing this text

```

&nbsp;
&nbsp;

```{r Importing Peer counties table}
county_fips <- read_excel("CHSIpeers.xlsx", sheet = 2) # reading the same excel file, sheet 2, to extract a dataframe
county_fips <- county_fips %>% #Renaming columns to remove blank spaces:
  transmute(FIPS = `County FIPS code`, County_Name_State = `County Name, State`, PeerCountyGroup = `Peer County Group`)

```

&nbsp;
&nbsp;

```{r Scraping names of counties and states}
county_fips <- county_fips %>% # using regex to separate county names and state names into two columns
  mutate(County = str_extract(County_Name_State, ".*(?=,)"),
          State = str_extract(County_Name_State, "(?<=, ).*")) %>% 
  select(-County_Name_State)

write_excel_csv(county_fips, "County_FIPS.csv") # writing into csv and excel files
write.xlsx(county_fips, "County_FIPS.xlsx")

```

Output files County_FIPS.xlsx and County_FIPS.csv are generated. County name and State names are put into two separate columns.

&nbsp;
&nbsp;

I am a resident of Passaic county, NJ. I will find what Strata my county belongs to, find Peer counties in this Strata, and count how many Peer counties are there:

```{r Passaic county Strata number, list of counties}
Passaic <- filter(county_fips,County == "Passaic County") # filtering out Passaic county line
Passaic <- as.character(Passaic$PeerCountyGroup)         # reading the strata number associated with it
PassaicStrata = str_c("Passaic county, NJ belongs to strata ", Passaic) # inserting this number into a chr string
print(PassaicStrata) # printing this chr string

PassaicStrata <- county_fips %>% filter(PeerCountyGroup == Passaic) # filtering out counties that belong to the same strata as Passaic
tc <- PassaicStrata %>% summarise(count = n()) # counting the number of observations
tc <- as.character(tc$count[[1]])             # extracting this number as a chr variable
# concatenating text and chr variables into one string: 
Mystrata_table_title = str_c("Peer counties in Strata ", Passaic, ", a total of ", tc, " counties")

```

\newpage

```{r strata 9 county listing}

kable(PassaicStrata, caption = Mystrata_table_title) # output of the county listing in strata 9

```




I will now work with the Demographics dataframe from the CHSI dataset, will look at the poverty levels by strata. Let's see how the strata my county belongs to compares to other stratas in the country.

```{r Poverty levels by strata}
demographics <- read.csv("DEMOGRAPHICS.csv") # importing the csv file

PovertyByStrata <- demographics %>% 
  group_by(Strata_ID_Number) %>%           # grouping by the strata id
  filter(Poverty > 0) %>%                 # removing lines that have negative values by filtering. Negative values stand for "no data available"
  summarise(Poverty = mean(Poverty))     # summarising the mean of Poverty for each strata

PovertyByStrata <- PovertyByStrata %>%
  arrange(Poverty) %>%                # arranging the lines by Poverty value with lowest on top
  mutate(rank = row_number())         # adding a rank column, assigning a rank by Poverty to each strata

StrataCount <- summarise(PovertyByStrata, count = n()) # counting how many stratas
StrataCount <- StrataCount[[1]]                        # assigning this value to a variable

```


```{r My strata ranking by Poverty level}

Strata9 <- PovertyByStrata %>% filter(Strata_ID_Number == Passaic) # filtering out the line for my strata

Strata9ranking <- Strata9$rank[[1]]      # assigning the rank value to a variable
Strata9pov <- Strata9$Poverty[[1]]      ## assigning the poverty value to a variable

# concatenating text and chr variables into one string and changing its class to a tibble: 
passaic_print <- as.tibble(str_c("Passaic county, NJ belongs to Strata ", Passaic, " that ranks ", Strata9ranking, "th lowest in Poverty level amongst the total of ", StrataCount, " Stratas in the US. \n", Strata9pov, "% of the population of Strata ", Passaic, " lives below poverty level."))

passaic_print <- passaic_print %>% rename(`...`=value)

kable(passaic_print, caption = "Passaic County, NJ") # printing the string above 

```

\newpage

Now that information on Strata 9 ranking is obtained, I will demonstrate it on a bar plot. Strata 9 is shown in blue, and the plot confirms the information received above. We see how most of the stratas have a higher poverty level.

&nbsp;
&nbsp;

```{r Bar plot of stratas by Poverty level}
# building a bar plot, showing strata 9 in blue:
ggplot(data = PovertyByStrata, aes(Strata_ID_Number, Poverty)) + 
  geom_bar(stat = "identity", alpha = .6) +
  geom_bar(data = Strata9, aes(Strata_ID_Number, Poverty, fill = Strata_ID_Number), stat = "identity", show.legend = TRUE)+ 
  ggtitle("Individuals living below poverty level per Strata, %") +
  xlab("Strata ID") + ylab("Poverty, %")

```

Let's glimpse at stratas with highest and lowest poverty levels:


```{r Strata with highest poverty level}
#arranging stratas by poverty in descending order to get an observation with the highest value
PovertyStrataMax <- PovertyByStrata %>% arrange(desc(Poverty)) %>% head(1)
psmax <- PovertyStrataMax[[1]]  # assigning this strata number to a variable.

PovertyMax_list <- demographics %>% 
  filter(Strata_ID_Number == psmax) %>% # filtering out lines that correspond to the strata number found above
  select(CHSI_County_Name, CHSI_State_Name, Poverty) %>% # leaving out columns we don't need
  arrange(desc(Poverty))                                # arranging by poverty in descending order

# preparing a header for a table to print:
psmax <- str_c("A strata with the highest % of population living below the poverty level is number ", psmax)
kable(head(PovertyMax_list), caption = psmax)  # printing this table

```




```{r Strata with lowest poverty level}
#arranging stratas by poverty in ascending order to get an observation with the lowest value
PovertyStrataMin <- PovertyByStrata %>% arrange(Poverty) %>% head(1)
psmin <- PovertyStrataMin[[1]]  # assigning this strata number to a variable.
  
PovertyMin_list <- demographics %>% 
  filter(Strata_ID_Number == psmin) %>% # filtering out by strata number found above
  select(CHSI_County_Name, CHSI_State_Name, Poverty) %>% 
  arrange(Poverty) #arranging by poverty in ascending order
#printing this table
kable(head(PovertyMin_list), caption = str_c("A strata with the lowest % of population living below the poverty level is number ", psmin))

```

&nbsp;
&nbsp;

The Population density data is provided by the Demographics file. I will select pertinent columns and clean the data of missing values.

```{r Extracting population density from demographics}
# Extracting columns for Population_Density.
# Negative values in Population_Density column stand for "no data available", these lines need to be removed
Pop_dens <- demographics %>% select(State_FIPS_Code, County_FIPS_Code, CHSI_County_Name, CHSI_State_Name, Strata_ID_Number, Population_Density) %>% 
  filter(Population_Density >= 0)

kable(head(select(Pop_dens, CHSI_County_Name, CHSI_State_Name, Population_Density)), caption = "County data, population density (people per square mile)")
```
&nbsp;
&nbsp;

 I will filter out and print 10 most populated counties in the country.

&nbsp;
&nbsp;

```{r Top 10 counties with highest population density}
#arranging counties by population density in descending order, selecting top 10
Top10Pop_dens <- Pop_dens %>% arrange(desc(Population_Density)) %>% head(10)
#printing this list
kable(select(Top10Pop_dens, CHSI_County_Name, CHSI_State_Name, Population_Density), caption = "Top 10 counties with highest population density in the US")

```


Four of the boroughs of New York City are leading the list leaving other counties far behind. No wonder they have the most of COVID-19 cases. Another proof that self isolation is essential to stop spreading COVID-19.

&nbsp;
&nbsp;

Now I will import and review the SUMMARYMEASURESOFHEALTH.csv file for the average life expectancy data. 

```{r average life expectancy from SummaryMeasuresOfHealth table}
#reading the csv file
SummaryMeasuresOfHealth <- read.csv("SUMMARYMEASURESOFHEALTH.csv")
# Extracting columns for average life expectancy (ALE).
# Negative values in ALE column stand for "no data available", these lines need to be removed
ALEdf <- SummaryMeasuresOfHealth %>% 
  select(State_FIPS_Code, County_FIPS_Code, CHSI_County_Name, CHSI_State_Name, Strata_ID_Number, ALE) %>% 
  filter(ALE > 0)
#print the head of the tibble:
kable(head(select(ALEdf, CHSI_County_Name, CHSI_State_Name, ALE)), caption = "County data, average life expectancy")

```
&nbsp;
&nbsp;
&nbsp;
&nbsp;

I will left join the Population density and the Life Expectancy data frames to produce a Linear Model of these two variables:

```{r join Pop_dens and ALEdf tables}
# left join of two df's and dropping any NA observations
Pop_dens_ALEdf <- drop_na(left_join(Pop_dens, ALEdf))

#print the head of the tibble:
kable(head(select(Pop_dens_ALEdf, CHSI_County_Name, CHSI_State_Name, Population_Density, ALE)), caption = "Population Density and Average Life Expectancy by county")

```
&nbsp;
&nbsp;
&nbsp;
&nbsp;

I built a Linear Model of ALE ~ Population_Density graph below, but does not look very informative. The scatter plot is jammed to the left. There are probably very few counties with high population density. I listed them previously in this report in Table 9.


```{r Linear Model ALE and Population_Density }
#linear model
ALElm <- lm(ALE ~ Population_Density, data=Pop_dens_ALEdf)

#extracting the list element containing fitted values and adding it to the original ALE and Population Density tibble
Pop_dens_ALEdf$fit <- ALElm[["fitted.values"]] 

```

```{r Plotting LM ALE and Pop dens}
# plotting the LM data obtained above. LM data is the red line, the original county data is a scatter plot in translucent blue
Pop_dens_ALEdf %>% 
  ggplot(aes(Population_Density, fit)) + geom_line(colour = 'red') + 
  ggtitle("Linear Model of ALE ~ Population_Density") +
  geom_point(aes(Population_Density, ALE), alpha = .3, color = "blue") + 
  ylab("Average Life Expectancy") +
  xlab("Population Density (people per sq mi)")


```



&nbsp;
&nbsp;
&nbsp;
&nbsp;
```{r LM ALE and Poverty}
Poverty_lvl <- demographics %>% 
  select(State_FIPS_Code, County_FIPS_Code, CHSI_County_Name, CHSI_State_Name, Strata_ID_Number, Poverty) %>% 
  filter(Poverty > 0)

ALE_Poverty_lvl <- drop_na(left_join(ALEdf, Poverty_lvl))

PovertyLM <- lm(ALE ~ Poverty, data=ALE_Poverty_lvl)

ALE_Poverty_lvl$fit <- PovertyLM[["fitted.values"]]

summary(ALE_Poverty_lvl)

```


&nbsp;
&nbsp;
&nbsp;
&nbsp;
```{r Plotting LM ALE and Poverty}
ALE_Poverty_lvl %>% 
  ggplot(aes(Poverty, fit)) + geom_line(colour = 'red') + 
  ggtitle("Linear Model of ALE ~ Poverty") +
  geom_point(aes(Poverty, ALE), alpha = .2, color = "blue") +
  ylab("Average Life Expectancy") +
  xlab("Individuals Living Below Poverty line, %")
```





&nbsp;
&nbsp;
&nbsp;
&nbsp;
```{r average life expectancy by county US map}
ALEdf1 <- ALEdf

ALEdf1$State_FIPS_Code <- as.character(ALEdf1$State_FIPS_Code)

ALEdf1$County_FIPS_Code <- as.character(ALEdf1$County_FIPS_Code)

ALEdf1$County_FIPS_Code <- if_else(str_length(ALEdf1$County_FIPS_Code) == 1, str_c("00", ALEdf1$County_FIPS_Code), ALEdf1$County_FIPS_Code)

ALEdf1$County_FIPS_Code <- if_else(str_length(ALEdf1$County_FIPS_Code) == 2, str_c("0", ALEdf1$County_FIPS_Code), ALEdf1$County_FIPS_Code)

ALEdf1$FIPS <- as.numeric(str_c(ALEdf1$State_FIPS_Code, ALEdf1$County_FIPS_Code))

ALEdf1 <- ALEdf1 %>% select(FIPS, CHSI_County_Name, CHSI_State_Name, ALE)

kable(head(ALEdf1), caption = "FIPS codes added to the Average Life Expectancy by county dataframe")

ALEdfPlot <- ALEdf1 %>% rename(value = ALE, region = FIPS)
county_choropleth(ALEdfPlot, title = "Average Life Expectancy by county",legend = "Age", num_colors = 9)


```

```{r average life expectancy in NJ and NY}
county_choropleth(ALEdfPlot, title = "New Jersey and New York, Average Life Expectancy by county",legend = "Age", state_zoom = c("new jersey", "new york"), num_colors = 9)
NJ_ALE <- county_choropleth(ALEdfPlot, title = "2010 NJ Average Life Expectancy",legend = "Age", state_zoom = "new jersey", num_colors = 9)
```


```{r exporting average life expectancy with FIPS codes in excel}
write.xlsx(ALEdf1, "ALE_with_FIPS.xlsx")
```

&nbsp;
&nbsp;
&nbsp;
&nbsp;

An excel file of average life expectancy with FIPS codes named ALE_with_FIPS.xlsx is created.




```{r New Jersey Rankings Data}

NJ2010 <- read_excel("2010 County Health Ranking New Jersey Data - v2.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2010" = `Adult obesity`)
NJ2010$"2010" <- as.double(NJ2010$"2010")

NJ2011 <- read_excel("2011 County Health Ranking New Jersey Data - v4.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2011" = `Adult obesity`)
NJ2011$"2011" <- as.double(NJ2011$"2011")

NJ2012 <- read_excel("2012 County Health Ranking New Jersey Data - v4.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2012" = `Adult obesity`)
NJ2012$"2012" <- as.double(NJ2012$"2012")

NJ2013 <- read_excel("2013 County Health Ranking New Jersey Data - v1_0.xls", sheet = 5) %>% 
  select(FIPS = `*Data for measures with an asterisk should not be compared with prior years due to changes in definition.`, County = `...3`, "2013" = `Adult obesity`)
NJ2013$"2013" <- as.double(NJ2013$"2013")

NJ2014 <- read_excel("2014 County Health Rankings New Jersey Data - v6.xls", sheet = 4) %>% 
  select(FIPS = `*Data for measures with an asterisk should not be compared with prior years due to changes in definition.`, County = `...3`, "2014" = `Adult obesity`)
NJ2014$"2014" <- as.double(NJ2014$"2014")

NJ2015 <- read_excel("2015 County Health Rankings New Jersey Data - v3.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2015" = `Adult obesity`)
NJ2015$"2015" <- as.double(NJ2015$"2015")

NJ2016 <- read_excel("2016 County Health Rankings New Jersey Data - v3.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2016" = `Adult obesity`)
NJ2016$"2016" <- as.double(NJ2016$"2016")

NJ2017 <- read_excel("2017 County Health Rankings New Jersey Data - v2.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2017" = `Adult obesity`)
NJ2017$"2017" <- as.double(NJ2017$"2017")

NJ2018 <- read_excel("2018 County Health Rankings New Jersey Data - v3.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2018" = `Adult obesity`)
NJ2018$"2018" <- as.double(NJ2018$"2018")

NJ2019 <- read_excel("2019 County Health Rankings New Jersey Data - v1_0.xls", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2019" = `Adult obesity`)
NJ2019$"2019" <- as.double(NJ2019$"2019")

NJ2020 <- read_excel("2020 County Health Rankings New Jersey Data - v1_0.xlsx", sheet = 4) %>% 
  select(FIPS = `...1`, County = `...3`, "2020" = `Adult obesity`)
NJ2020$"2020" <- as.double(NJ2020$"2020")

NJobesity <- left_join(NJ2010, NJ2011)
NJobesity <- left_join(NJobesity, NJ2012)
NJobesity <- left_join(NJobesity, NJ2013)
NJobesity <- left_join(NJobesity, NJ2014)
NJobesity <- left_join(NJobesity, NJ2015)
NJobesity <- left_join(NJobesity, NJ2016)
NJobesity <- left_join(NJobesity, NJ2017)
NJobesity <- left_join(NJobesity, NJ2018)
NJobesity <- left_join(NJobesity, NJ2019)
NJobesity <- drop_na(left_join(NJobesity, NJ2020))

kable(NJobesity, caption = "New Jersey Adult Obesity levels, %")

NJ2020 <- NJ2020 %>% rename(value = "2020", region = FIPS) %>% drop_na()
NJ2020$region <- as.numeric(NJ2020$region)

NJobesity_plot <- county_choropleth(NJ2020, title = "2020 NJ Obesity levels by Year, %", state_zoom = "new jersey", num_colors = 9)

library(gridExtra)
grid.arrange(NJ_ALE, NJobesity_plot, ncol=2)



```


```{r Obesity rates by year}

NJobesity1 <- NJobesity %>% 
  gather(Year, obesity, 3:13) %>% 
  group_by(Year) %>% 
  summarise(obesity = mean(obesity))
NJobesity1$Year <- as.double(NJobesity1$Year)

ggplot(NJobesity1, aes(Year, obesity)) + 
  geom_point(aes(size = obesity^15), color = "darkred", show.legend = FALSE) + 
  geom_line() + geom_smooth() +
  ggtitle("New Jersey Adult Obesity Rates") +
  xlab("Year") + ylab("Obesity, % of population")

```



&nbsp;
&nbsp;
&nbsp;
&nbsp;

\newpage

Bibliography:

1. The County Health Rankings & Roadmaps program. (August 30, 2017). Peer Counties Tool.

Retrieved 30 April 2020, from https://www.countyhealthrankings.org/resources/peer-counties-tool

2. The County Health Rankings & Roadmaps program. (2020). New Jersey Rankings Data.

Retrieved 30 April 2020, from https://www.countyhealthrankings.org/app/new-jersey/2020/downloads

3. U.S. Government’s open data. (February 26, 2020). Community Health Status Indicators (CHSI) to Combat Obesity, Heart Disease and Cancer.

Retrieved 30 April 2020, from https://catalog.data.gov/dataset/community-health-status-indicators-chsi-to-combat-obesity-heart-disease-and-cancer
