---
title: "Final project"
author: "Elena Novikova"
date: "4/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



```{r}

library(tidyverse)
library(lubridate)
library(magrittr)
library(readxl)
library(dplyr)
library(stringr)
library(openxlsx)
library(knitr)
library(rvest)

```

```{r Web scraping from data.gov}

CHSIdescrip <- "https://catalog.data.gov/dataset/community-health-status-indicators-chsi-to-combat-obesity-heart-disease-and-cancer" %>%
  read_html() %>% html_nodes(".module-content") %>%
  html_text()
CHSIdescrip <- tibble(text = CHSIdescrip)
CHSIdescrip <- CHSIdescrip[[1,1]]
#strwrap(CHSIdescrip, width = 80)


CHSIheader <- str_extract(CHSIdescrip, "[^\n \t].*")
CHSIdate <- str_extract(CHSIdescrip, "M.*")
CHSItext <- str_extract(CHSIdescrip, "\n +C.*") 
CHSItext <- str_remove_all(CHSItext, "\n  +")
CHSItable <- tibble(x = c(CHSIheader, CHSIdate, CHSItext))
kable(CHSItable, caption = "CHSItable")
```


```{r Peer counties decription}
county_fips_descrip <- read_excel("CHSIpeers.xlsx", sheet = 1)
county_fips_descrip <- county_fips_descrip[[2,1]]
kable(county_fips_descrip, caption = "Peer Counties")

```


```{r Importing Peer counties table}
county_fips <- read_excel("CHSIpeers.xlsx", sheet = 2)
#Renaming columns to remove blank spaces:
county_fips <- county_fips %>% 
  transmute(FIPS = `County FIPS code`, County_Name_State = `County Name, State`, PeerCountyGroup = `Peer County Group`)

county_fips$FIPS <- as.character(county_fips$FIPS)
# leading zeros are missing from FIPS codes after importing from excel. Adding this zero:
county_fips$FIPS <- if_else(str_length(county_fips$FIPS) == 4, str_c("0", county_fips$FIPS), county_fips$FIPS)

```

```{r Scraping FIPS codes}
county_fips <- county_fips %>% 
  mutate(County = str_extract(County_Name_State, ".*(?=,)"),
          State = str_extract(County_Name_State, "(?<=, ).*")) %>% 
  select(-County_Name_State)

write_excel_csv(county_fips, "County_FIPS.csv")
write.xlsx(county_fips, "County_FIPS.xlsx")

```

I am a resident of Passaic county, NJ. Let's find what Strata my county belongs to, find Peer counties in this Strata, and count how many Peer counties are there:

```{r Passaic county Strata}
x <- filter(county_fips,County == "Passaic County")
x <- as.character(x$PeerCountyGroup)
y = str_c("Passaic county, NJ belongs to strata ", x)
print(y)

My_strata <- county_fips %>% filter(PeerCountyGroup == x)
tc <- My_strata %>% summarise(count = n())
tc <- as.character(tc$count[[1]])

Mystrata_table_title = str_c("Peer counties in Strata ", x, ", a total of ", tc, " counties")

kable(My_strata, caption = Mystrata_table_title)


```

```{r Poverty levels by strata}
demographics <- read.csv("DEMOGRAPHICS.csv")


PovertyByStrata <- demographics %>% 
  group_by(Strata_ID_Number) %>%
  filter(!str_detect(Poverty, "-")) %>% 
  summarise(Poverty = mean(Poverty))

strata37 <- demographics %>% 
  filter(Strata_ID_Number == 37) %>% 
  select(CHSI_County_Name, CHSI_State_Name, Poverty)

Strata9 <- PovertyByStrata %>% 
  filter(Strata_ID_Number == 9)

ggplot(data = PovertyByStrata, aes(Strata_ID_Number, Poverty)) +
  geom_bar(stat = "identity") +
  geom_bar(data = Strata9, aes(Strata_ID_Number, Poverty, fill = Strata_ID_Number), stat = "identity", show.legend = FALSE)+ 
  ggtitle("County data per strata, individuals living below poverty level, %") +
  xlab("Strata ID") + ylab("Poverty, %")
```


```{r Extracting population density from demographics}
# Extracting columns for Population_Density.
# Negative values in Population_Density column stand for "no data available", these lines need to be removed
Pop_dens <- demographics %>% select(State_FIPS_Code, County_FIPS_Code, CHSI_County_Name, CHSI_State_Name, Strata_ID_Number, Population_Density) %>% 
  filter(Population_Density >= 0)
```

```{r average life expectancy from SummaryMeasuresOfHealth table}
SummaryMeasuresOfHealth <- read.csv("SUMMARYMEASURESOFHEALTH.csv")
# Extracting columns for average life expectancy (ALE).
# Negative values in ALE column stand for "no data available", these lines need to be removed
ALEdf <- SummaryMeasuresOfHealth %>% 
  select(State_FIPS_Code, County_FIPS_Code, CHSI_County_Name, CHSI_State_Name, Strata_ID_Number, ALE) %>% 
  filter(ALE > 0)


```

```{r join Pop_dens and ALEdf tables}

```





Bibliography:

1. The County Health Rankings & Roadmaps program. (August 30, 2017). Peer Counties Tool.

Retrieved 30 April 2020, from https://www.countyhealthrankings.org/resources/peer-counties-tool

2. The County Health Rankings & Roadmaps program. (2020). New Jersey Rankings Data.

Retrieved 30 April 2020, from https://www.countyhealthrankings.org/app/new-jersey/2020/downloads

3. U.S. Government’s open data. (February 26, 2020). Community Health Status Indicators (CHSI) to Combat Obesity, Heart Disease and Cancer.

Retrieved 30 April 2020, from https://catalog.data.gov/dataset/community-health-status-indicators-chsi-to-combat-obesity-heart-disease-and-cancer
